Validation of Visual Statistical Inference, Applied to Linear Models

### 2012-07-10 paper_jasa.tex
finished revising the p-value difference, conclusions, and figure captions

The model in Equation \ref{eqn:mixed} is fit using $E$ as the only fixed effect covariate without intercept. we set offset = $log(0.05/0.95)$ so that the estimated power has a fixed lower limit at 0.05 (Type-I error) when $E=0$. Different subjects are expected to have different skill levels in evaluating lineups. To account for subject specific variation we use subject as random effect. We do not include random intercept, instead, we use an interaction between subject and effect $E$ to allow for subject specific random slope. This leads to setting  $W_{\ell i} = E_{\ell i}$, the effect of lineup $\ell$ that subject $i$ has evaluated in Equation \ref{eqn:mixed}.  While for experiment 1 and 2 we don't allow intercept but offset, for experiment 3 an intercept is estimated to allow the power curves to go above 0.05 when $E=0$. This is because forcing power to be fixed at 0.05 for $E=0$ is not justified by the experiment 3 data (Figure \ref{fig:power_loess_effect}).

@@

Table \ref{tbl:model_par} shows the parameter estimates of the mixed effects model of the subject-specific variation. The larger variability in the response data for experiment 3 as we see in Figure \ref{fig:power_loess_effect} contributes the random effect variance estimate be much larger compared to those of other experiments. Because fo the same reason we see the standard error for intercept estimate is much higher as well. Also notice the slope estimate for experiment three is lowest as well. It is because for small parameter values the power is much higher for experiment 3.  % \green{What do we learn about the models from these parameter estimates? Explain this here too.}

@@

Figures \ref{fig:P-val_log} and \ref{fig:P-val_log2} show the selection counts and $p$-values of all plots for each lineup in experiments 1 and 2, respectively. In each of these plots, the log of $p$-value for each plot in the lineup is plotted horizontally and the count of the number of subjects who selected this plot is displayed vertically. The dot represents this point, and the lines connect vertically with 0 count. Each row corresponds to a different $\beta$ value ordered top to bottom from lowest to highest. Columns correspond to the other levels of the experiment, sample size $(n)$, error standard deviation $(\sigma)$ and replicate. (Empty cells indicate no lineup shown with this combination.) Red indicates the plot with the lowest $p$-value in the lineup. In both experiments as $\beta$ gets larger (stronger effect) people tended to select the plot with the lowest $p$-value. This is less so the case when $\beta$ is small, when standard deviation is large and sample size is small. The results are clearer for experiment 2, that used a continuous covariate. But even when they didn't pick the plot with the lowest $p$-value they tended to oscillate their choices between the several low $p$-value plots. So for most subjects, the assumption that they pick the plot with the smallest $p$-value would appear to be reasonable, and the actual power of the visual test should be close to the expected power.

\begin{figure*}[hbtp]
   \centering
       \includegraphics[width=0.95\textwidth]{p_val_log_counts.pdf}
       \caption{Relative frequency of plot picks compared to other plots in the lineup plotted against the $p$-value (on log$_{10}$ scale) of each plot for all of the sixty (60) individual lineups for experiment 1. The colored lines indicate the plot with the lowest $p$-value. Rows correspond to $\beta$, smallest to largest values from top to bottom. Columns correspond to other experimental treatments, sample size $(n)$, error standard deviation $(\sigma)$, and replicate. Empty cells indicate no lineup for that combination of parameters. Highest counts tend to be the plot in the lineup having the lowest $p$-value.}
       \label{fig:P-val_log}
\end{figure*}

\begin{figure*}[hbtp]
   \centering
       \includegraphics[width=0.95\textwidth]{p_val_log_counts2.pdf}
       \caption{Relative frequency of plot picks compared to other plots in the lineup plotted against the $p$-value (on log$_{10}$ scale) of each plot for all of the seventy (70) individual lineups for experiment 2. The colored lines indicate the plot with the lowest $p$-value. Rows correspond to $\beta$, smallest to largest values from top to bottom. Columns correspond to other experimental treatments, sample size $(n)$, error standard deviation $(\sigma)$, and replicate. Empty cells indicate no lineup for that combination of parameters. Highest counts tend to be the plot in the lineup having the lowest $p$-value, even more so than for experiment 1.}
       \label{fig:P-val_log2}
\end{figure*}

@@

Visual inference is like other types of computationally intensive test procedures, such as permutation tests, where the statistics from the data are compared with those from null data. But unlike these approaches visual inference is constrained to make the comparison with just a few draws $(m-1)$ from the null distribution. How this small set of null plots influences the subjects choice is important for understanding the reliability of visual inference. If the actual data plot is very different from the null plots, then the null plots should not have much influence on the choice. Measuring the difference between the plots is almost impossible! But in the controlled setting here we can again use $p$-value as a proxy for similarity and difference.

In the previous section we noticed that subjects tend to pick the plot with the lowest $p$-value. If a null plot has data that, by chance because the sample is a random draw from the null, has a smaller $p$-value than the actual data, we would expect that there is a strong possibility that an subject will pick this null plot instead of the actual data plot. Figure \ref{fig:pval_difference} investigates this for the data collected in experiments 1 and 2. The plots display the proportion of subjects who choose the actual data plot against the $p$-value of the actual data minus the lowest seen in the null data. A small negative value of the difference means that the $p$-value of the actual data is smaller than that of the null data. If there is a big negative difference we'd expect to see that the subjects would more often make the correct choice of plot. When there is a large positive difference, subjects fail to pick the correct plot. If however, there is little difference between the $p$-values then we'd expect to see more confusion in the choices. That is why we see a big range of proportion correct where difference is close to zero in Figure \ref{fig:pval_difference}.

@@

\begin{figure}[hbtp]
   \centering
       \scalebox{0.70}{\includegraphics{pval_difference.pdf}}
       \caption{Difference between the data plot's $p$-value and the  smallest $p$ value of a null plot vs proportion correct.}
       \label{fig:pval_difference}
\end{figure}

@@

It is widely accepted that statistical graphics has important use in exploratory data analysis and model diagnostics. This paper has demonstrated that statistical graphics can be used in statistical inference and validates the lineup protocol proposed by \citet{buja:2009} as a formal framework to accomplish that. The simulation study showed that even in the situation where the conventional test has its assumptions met, visual inference is comparable. It is possible that, for some trained observers visual inference may produce even better power. The results also suggest that when the underlying assumptions of a conventional test do not hold, visual inference will perform well.

@@

\blue{Another critical part is the  question that the observer needs to answer when evaluating the lineup: this question can be lined up with the alternative hypothesis of a classical testing scenario. 
If we are interested in the reason why participants picked a plot as the most different, we need to collect and evaluate this information in the study. This has to be done very carefully, as any suggestions for reasons also lead the observer  what patterns to look out for.}
% By delivering an appropriate question may reduce the type III error in an appropriate scenario.}

@@

No restrictions were placed on Turk workers, in terms of abilities. It would be interesting to see how well the trained subjects performs. Many pilot studies with advanced knowledge of statistical graphics as well as statistics indicates that the performance is very good for trained subjects. One of our future work would be to examine this in details. For this we are planning to trained Turk subjects with specific graphics and educate them what pattern a specific graphics may show. After being trained they may have to pass the qualification test before participating the experiment. Even though we did not do this, it is very common to Turk subjects to go through this process for various other tasks they do in the Turk web site.

>>>>

The model in Equation \ref{eqn:mixed} is fit using $E$ as the only fixed effect covariate without intercept. we set offset = $log(0.05/0.95)$ so that the estimated power has a fixed lower limit at 0.05 (Type-I error) when $E=0$. Different subjects are expected to have different skill levels in evaluating lineups. To account for subject specific variation we use subject as random effect. We do not include random intercept, instead, we use an interaction between subject and effect $E$ to allow for subject specific random slope. This leads to setting  $W_{\ell i} = E_{\ell i}$, the effect of lineup $\ell$ that subject $i$ has evaluated in Equation \ref{eqn:mixed}.  While for experiment 1 and 2 we don't allow intercept but offset, for experiment 3 an intercept is estimated to allow the power curves to go above 0.05 when $E=0$. This is because forcing power to be fixed at 0.05 for $E=0$ is not justified by the experiment 3 data.% (Figure \ref{fig:power_loess_effect}). *** You cannot reference the figure before you explain it!!!

@@

Table \ref{tbl:model_par} shows the parameter estimates of the mixed effects model of the subject-specific variation. \green{The fixed effects estimates indicate that for all experiments as the effect increases the proportion correct increases. For experiment 3 it is very weak. The subject-specific variability is smaller for experiment 1, and relatively large for experiment 3.} %The larger variability in the response data for experiment 3% as we see in Figure \ref{fig:power_loess_effect}  *** You cannot reference the figure before you explain it!!!
%contributes to the random effect variance estimate be much larger compared to those of other experiments. For the same reason we see the standard error for intercept estimate is much higher as well. Also notice the slope estimate for experiment three is lowest as well. It is because for small parameter values the power is much higher for experiment 3.  % \green{What do we learn about the models from these parameter estimates? Explain this here too.}

@@

Figures \ref{fig:P-val_log} and \ref{fig:P-val_log2} show the selection counts and $p$-values of all plots for each lineup in experiments 1 and 2, respectively. In each of these plots, the log of $p$-value for each plot in the lineup is plotted horizontally and the count of the number of subjects who selected this plot is displayed vertically. The dot represents this point, and the lines connect vertically with 0 count. Each row corresponds to a different $\beta$ value ordered top to bottom from lowest to highest. Columns correspond to the other levels of the experiment, sample size $(n)$, error standard deviation $(\sigma)$ and replicate. (Empty cells indicate no lineup shown with this combination.) Red indicates the plot with the lowest $p$-value in the lineup. (Blue indicates the plot of the actual data when it is different from that with the lowest $p$-value.) In both experiments as $\beta$ gets larger (stronger effect) people tended to select the plot with the lowest $p$-value. This is less so the case when $\beta$ is small, when standard deviation is large and sample size is small. The results are clearer for experiment 2, that used a continuous covariate. But even when they didn't pick the plot with the lowest $p$-value they tended to oscillate their choices between the several low $p$-value plots. So for most subjects, the assumption that they pick the plot with the smallest $p$-value would appear to be reasonable, and the actual power of the visual test should be close to the expected power.

\begin{figure*}[hbtp]
   \centering
       \includegraphics[width=0.95\textwidth]{p_val_log_counts.pdf}
       \caption{Relative frequency of plot picks compared to other plots in the lineup plotted against the $p$-value (on log$_{10}$ scale) of each plot for all of the sixty (60) individual lineups for experiment 1. Red indicates the plot with the lowest $p$-value, and blue indicates the actual data plot, when it is different from that with the lowest $p$-value. Rows correspond to $\beta$, smallest to largest values from top to bottom. Columns correspond to other experimental treatments, sample size $(n)$, error standard deviation $(\sigma)$, and replicate. Empty cells indicate no lineup for that combination of parameters. Highest counts tend to be the plot in the lineup having the lowest $p$-value.}
       \label{fig:P-val_log}
\end{figure*}

\begin{figure*}[hbtp]
   \centering
       \includegraphics[width=0.95\textwidth]{p_val_log_counts2.pdf}
       \caption{Relative frequency of plot picks compared to other plots in the lineup plotted against the $p$-value (on log$_{10}$ scale) of each plot for all of the seventy (70) individual lineups for experiment 2. Red indicates the plot with the lowest $p$-value, and blue indicates the actual data plot, when it is different from that with the lowest $p$-value. Rows correspond to $\beta$, smallest to largest values from top to bottom. Columns correspond to other experimental treatments, sample size $(n)$, error standard deviation $(\sigma)$, and replicate. Empty cells indicate no lineup for that combination of parameters. Highest counts tend to be the plot in the lineup having the lowest $p$-value, even more so than for experiment 1.}
       \label{fig:P-val_log2}
\end{figure*}

@@

Visual inference is like other types of computationally intensive test procedures, such as permutation tests, where the statistics from the data are compared with those from null data. But unlike these approaches visual inference is constrained to make the comparison with just a few draws $(m-1)$ from the null distribution. How this small set of null plots influences the subjects choice is important for understanding the reliability of visual inference. If the actual data plot is very different from the null plots, then the null plots should not have much influence on the choice. Measuring the difference, generally, between the plots is almost impossible! But in the controlled setting here we can use $p$-value of the test statistic calculated on the data used in each plot as a proxy for similarity of structure between the plots.

We have seen that the subjects tend to pick the plot in the lineup that has the lowest $p$-value (Figures \ref{fig:P-val_log}, \ref{fig:P-val_log2}). What we are interested in here is how this pick is affected by the distributin=on of $p$-values of other plots in the lineup, particularly the $p$-value of the null plot with the strongest structure. If there is a null plot with a small $p$-value, or one close to that of the actual data plot, we would expect that subjects have a harder time detecting the actual data plot. Figure \ref{fig:pval_difference} investigates this. The difference between the $p$-value of the actual data is compared with the lowest from the null plots. This is plotted horizontally, and the proportion correct is plotted vertically. Negative values indicate lineups where the actual data plot had a smaller $p$-value than the minimum of the null plots. In experiment 1 (boxplots) there were a lot more lineups where the actual data plot had the smaller $p$-value, but only barely. This caused quite some confusion for subjects, as seen because the variability in the proportion correct is huge for these lineups. Similarly large variability in correctness can be seen in the results of experiment 2 (scatterplots) except that the greater range of of differences in $p$-values shows the strength of subject's ability to pick the plot with most structure. Figures \ref{fig:P-val_log}, \ref{fig:P-val_log2} show a little more complicated story. When there is a big difference between the $p$-values (eg experiment 1, $\beta>7$) the subjects' as a force choose the same plot. when there is less difference the distribution of counts is much more evenly spread between plots (eg experiment 1, $\beta=1$).

%\blue{In the previous section we noticed that subjects tend to pick the plot with the lowest $p$-value. If a null plot has data that, by chance because the sample is a random draw from the null, has a smaller $p$-value than the actual data, we would expect that there is a strong possibility that an subject will pick this null plot instead of the actual data plot. Figure \ref{fig:pval_difference} investigates this for the data collected in experiments 1 and 2. The plots display the proportion of subjects who choose the actual data plot against the $p$-value of the actual data minus the lowest seen in the null data. A small negative value of the difference means that the $p$-value of the actual data is smaller than that of the null data. If there is a big negative difference we'd expect to see that the subjects would more often make the correct choice of plot. When there is a large positive difference, subjects fail to pick the correct plot. If however, there is little difference between the $p$-values then we'd expect to see more confusion in the choices. That is why we see a big range of proportion correct where difference is close to zero in Figure \ref{fig:pval_difference}.}

@@

\begin{figure}[hbtp]
   \centering
       \scalebox{0.70}{\includegraphics{pval_difference.pdf}}
       \caption{Investigating how does the strength of structure in the null plots affects the choice of the actual data plot. Difference between the data plot's $p$-value and the  smallest $p$ value of a null plot vs proportion correct. Negative values indicate the $p$-value of the actual data plot are smaller than those of all of the null plots. Close to zero shows a wide range in the proportion correct, suggesting that when at least one null plot has structure almost as strong as the actual data plot, subjects had a difficult time in making their choice.}
       \label{fig:pval_difference}
\end{figure}

@@

It is widely accepted that statistical graphics has important use in exploratory data analysis and model diagnostics. This paper has demonstrated that statistical graphics can be used in statistical inference and validates the lineup protocol proposed by \citet{buja:2009} as a formal framework to accomplish that. The simulation study showed that even in the situation where the conventional test has its assumptions met, visual inference is comparable. The results also suggest that when the underlying assumptions of a conventional test do not hold, visual inference will perform well.

@@

The question that is posed to observers can affect the abililty to evaluate lineups. In these experiments subjects were asked very specific questions, because the purpose was to compare with very specific conventional testing. In practice, the question posed will be very general to allow observers to choose plots with any remarkable features, which enables the data mining discovery process.

% when evaluating the lineup: this question can be lined up with the alternative hypothesis of a classical testing scenario. If we are interested in the reason why participants picked a plot as the most different, we need to collect and evaluate this information in the study. This has to be done very carefully, as any suggestions for reasons also lead the observer  what patterns to look out for.}
% By delivering an appropriate question may reduce the type III error in an appropriate scenario.}

@@

The performance of subjects was quite varied, but consistent. No restrictions were placed on Turk workers, in terms of abilities. There were clearly some subjects who performed very badly, but it was very interesting to see that there were some super-observers, people who detected the actual data plot at a rate better than that of the power of the best conventional test. 
 It would be interesting to see how well trained subjects might perform. Prior to the Turk experiments, we conducted pilot studies using local graphics experts and obtained good results, indicating that training in data visualization might be helpful for visual inference. Future work might explore this.
